---
title: "NYPD shooting"
output: 
  html_document: 
    highlight: tango
    theme: cerulean
    df_print: kable
date: "`r Sys.Date()`"
---

### Dataset used for this analysis

-   NYPD Shooting Incident Data available on DATA.GOV

-   Dataset used is a list of every shooting incident that occurred in NYC from 2006 to 2022

-   Each record represents a shooting incident in NYC and includes information about the event, the location and time of occurrence, and information related to suspects and victim demographics.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
library(tidyverse)
```

```{r}
# get data

url_in <-"https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

shooting_raw <- read_csv(url_in)

```

```{r}
# tidy data

shooting <- shooting_raw %>% 
  mutate(OCCUR_DATE = mdy(OCCUR_DATE),
         BORO = as_factor(BORO),
         PERP_AGE_GROUP = as_factor(PERP_AGE_GROUP),
         PERP_SEX = as_factor(PERP_SEX),
         PERP_RACE = as_factor(PERP_RACE),
         VIC_AGE_GROUP = as_factor(VIC_AGE_GROUP),
         VIC_SEX = as_factor(VIC_SEX),
         VIC_RACE = as_factor(VIC_RACE),
         occur_year = year(OCCUR_DATE),
         occur_month = factor(format(OCCUR_DATE, "%b"), levels = month.abb),
         occur_hour = hour(OCCUR_TIME)) %>% 
  janitor::clean_names()
```

#### Let's look at the summary of data first.

```{r}
summary(shooting)

shooting %>% skimr::skim()
```

#### How many shooting incidents and deaths by year?

```{r}
shooting %>% 
  group_by(occur_year) %>% 
  summarise(cases = n(), deaths = sum(statistical_murder_flag)) %>% 
  ggplot(aes(x = occur_year, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  expand_limits(y = 0) +
  theme_minimal() +
  labs(title = "New York Shooting Incidents by Year",
       x = "Year",
       y = "Cases")
```


You can see that cases were declining till 2019 but it came back up in 2020 and after. There is no significant change in the number of deaths.

<br><br>


#### How many shooting incidents and deaths by year by borough?
```{r}
shooting %>% 
  group_by(occur_year, boro) %>% 
  summarise(cases = n(), deaths = sum(statistical_murder_flag)) %>% 
  ggplot(aes(x = occur_year, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  expand_limits(y = 0) +
  facet_wrap(~boro) +
  theme_minimal() +
  labs(title = "Shooting Incidents Trend by Borough (2006-2022)",
       x = "Year",
       y = "Cases")


```


Brooklyn overall has the most shooting incidents and Staten Island has the least.


<br><br>


#### Is there a seasonality trend?
```{r}
shooting %>% 
  count(occur_month) %>% 
  ggplot(aes(occur_month, n)) +
  geom_col(fill = "#AFEEEE") +
  theme_minimal() +
  labs(title = "Shooting Incidents per Month",
       x = "Month",
       y = "Cases")

shooting %>% 
  count(occur_month, occur_year) %>% 
  ggplot(aes(occur_month, n)) +
  geom_col(fill = "#AFEEEE") +
  labs(title = "Seasonality Trend by Year",
       x = "Month",
       y = "Cases") +
  facet_wrap(~occur_year) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```


There is an increase over the summer months. And to make sure there is no outliers I broke it down by year. Here you can see the seasonality pattern exists every year.


<br><br>

#### Is there any pattern for time of day?
```{r}
shooting %>% 
  group_by(occur_hour) %>% 
  summarise(cases = n(), deaths = sum(statistical_murder_flag)) %>% 
  ggplot(aes(x = occur_hour, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  expand_limits(y = 0) +
  theme_minimal() +
  labs(title = "Shooting Incidents by Hour",
       x = "Hour",
       y = "Cases")
```



The number of cases significantly decreases in the morning and during daytime. And it increases in the evening and highest around midnight.


<br><br>

```{r}

# Fit a logistic regression model
model <- glm(statistical_murder_flag ~ vic_sex + vic_race, 
             data = shooting, 
             family = "binomial")

# Summary of the model
summary(model)
```

```{r}

# Create a dataframe with data for prediction
prediction_data <- expand.grid(
  vic_sex = levels(shooting$vic_sex),
  vic_race = levels(shooting$vic_race)
)

# Predict probabilities using the fitted model
prediction_data$predicted_prob <- predict(model, newdata = prediction_data, type = "response")

# Create a bar plot of predicted probabilities
prediction_data %>% 
ggplot(aes(x = interaction(vic_sex, vic_race), y = predicted_prob)) +
  geom_bar(stat = "identity", fill = "#AFEEEE") +
  labs(title = "Predicted Probabilities of Murder Cases",
       x = "Group",
       y = "Predicted Probability") +
  theme_minimal() +
  coord_flip()
```



In this logistic regression model, Pacific Islander and White victims have a higher predicted probabilities for murder cases.
